---
title: "Project STATS"
output:
  word_document: default
  pdf_document: default
date: "2024-04-27"
---


```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tidyr)
```

```{r}
data <- read.csv ("C:\\Users\\Dr Lavanya\\OneDrive\\Desktop\\Lavanya\\assignment\\applied stats\\project\\kidney_disease.csv")
View(data)
data= subset(data, select = -c(X))
data
```

#To examine the dimensions and the structure of the data
```{r}
#To examine the dimensions and the structure of the data
dim (data)

str (data)
```
```{r}
#Checking for null values in the entire data frame
if (any (is.na(data))) {
  print ("There are NA values in the data frame")
} else {
  print ("There are no NA values in the data frame")
}
```

```{r}
#Checking for null values
total_na <- sum (is.na (data))
total_na
```

```{r}
any (duplicated (data))
```

```{r}

####################################################### UPDATED CODE ##################################################################
na_count_by_column <- sapply(data, function(x) sum (is.na(x)))
na_count_by_column


#Imputing the missing values in numerical data with median
# Assuming 'data' is your dataframe
data$Blood.Pressure[is.na(data$Blood.Pressure)] <- median(data$Blood.Pressure, na.rm = TRUE) 
data$Serum.creatinine[is.na(data$Serum.creatinine)] <- median(data$Serum.creatinine, na.rm = TRUE)
data$Blood.urea[is.na(data$Blood.urea)] <- median(data$Blood.urea, na.rm = TRUE)  # For numerical data
data$Albumin[is.na(data$Albumin)] <- median(data$Albumin, na.rm = TRUE)  # For numerical data
data$Sugar[is.na(data$Sugar)] <- median(data$Sugar, na.rm = TRUE)  # For numerical data
data$Specific.Gravity[is.na(data$Specific.Gravity)] <- median(data$Specific.Gravity, na.rm = TRUE) # For numerical data
data$Blood.glucose.random[is.na(data$Blood.glucose.random)] <- median(data$Blood.glucose.random, na.rm = TRUE)  # For numerical data
data$Sodium[is.na(data$Sodium)] <- median(data$Sodium, na.rm = TRUE)  # For numerical data
data$Potassium[is.na(data$Potassium)] <- median(data$Potassium, na.rm = TRUE)  # For numerical data
data$Hemoglobin[is.na(data$Hemoglobin)] <- median(data$Hemoglobin, na.rm = TRUE)  # For numerical data
data$Packed.cell.volume[is.na(data$Packed.cell.volume)] <- median(data$Packed.cell.volume, na.rm = TRUE)
data$White.blood.cell.count[is.na(data$White.blood.cell.count)] <- median(data$White.blood.cell.count, na.rm = TRUE)
data$Red.blood.cell.count[is.na(data$Red.blood.cell.count)] <- median(data$Red.blood.cell.count, na.rm = TRUE)
view(data)


data$Age[is.na(data$Age)] <- median(data$Age, na.rm = TRUE)
median(data$Age, na.rm = TRUE)


# Replaced the missing values in numerical data with Median


na_count_by_column <- sapply(data, function(x) sum (is.na(x)))
na_count_by_column
view(data)


############### FILLING "" VALUES IN CATEGORICAL COLUMNS
# Function to calculate mode
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Identify categorical columns
cat_columns <- sapply(data, is.character)

# Loop through each categorical column
for (col in names(data)[cat_columns]) {
  # Calculate mode of the column
  mode_val <- get_mode(data[[col]])
  
  # Replace empty strings with mode value
  data[[col]][data[[col]] == ""] <- mode_val
}


########################### Since column "X.1" has unique values : ["#N/A", "#NAME?", ""], we should drop the column.
########################### Values doesn't convey any information. 
data_updated <- subset(data, select = -"$x.1")
view(data)
```
```{r}
#data_updated = data[, -which(names(data)== "X.1", "X")]
data_updated = data[, !(names(data) %in% c("X.1", "X"))]
view(data_updated)
```

Summary stats
```{r}
summary(data)
summary_data <- do.call(cbind, data_updated)
```

Calculating Range, Skewness
```{r}
# Function to calculate range
range_data_updated <- function(x) {
  if(is.numeric(x)) {
    range_val <- max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
    return(range_val)
  } else {
    return(NA)  # return NA for non-numeric variables
  }
}

# Function to calculate skewness
skewness_data_updated <- function(x) {
  if(is.numeric(x)) {
    skewness_val <- (3 * (mean(x, na.rm = TRUE) - median(x, na.rm = TRUE))) / sd(x, na.rm = TRUE)
    return(skewness_val)
  } else {
    return(NA)  # return NA for non-numeric variables
  }
}

# Applying functions to each variable
result <- data.frame(
  Variable = names(data_updated),
  Range = sapply(data_updated, range_data_updated),
  Skewness = sapply(data_updated, skewness_data_updated)
)
result
```

Check outliers and IQR
```{r}
# Checking for the presence of outliers
for (col in names(data_updated)[sapply(data_updated, is.numeric)]) {
  print(paste("Column:", col))
  
  # Calculate IQR
  q1 <- quantile(data_updated[[col]], 0.25, na.rm = TRUE)
  q3 <- quantile(data_updated[[col]], 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  print(paste("IQR:", iqr))
  
  # Identify outliers
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  outliers <- data_updated[[col]][data_updated[[col]] < lower_bound | data_updated[[col]] > upper_bound]
  print(paste("Outliers:", length(outliers)))
  print(outliers)
  
  #print("---")
}
```

Bar plots for categorical variables

```{r}
ggplot(data_updated, aes(x = hypertension, fill = hypertension)) + geom_bar() + scale_fill_manual(values = c("yes" = "red", "no" = "green"))
ggplot(data_updated, aes(x = Diabetes.Mellitus, fill = Diabetes.Mellitus)) + geom_bar() + scale_fill_manual(values = c("yes" = "orange", "no" = "blue"))
# Using scale_fill_manual to specify custom colors
ggplot(data_updated, aes(x = as.factor(Classification), fill = as.factor(Classification))) +
  geom_bar() +
  scale_fill_manual(values = c("ckd" = "pink", "notckd" = "skyblue")) +  # Adjusted color mapping
  labs(
    x = "Classification",
    y = "Count",
    fill = "CKD Classification",
    title = "Bar Chart of CKD Classification"
  )
```

Histogram for age vs classification of the disease

```{r}
ggplot(data_updated, aes(x = Age, fill = as.factor(Classification))) +
  geom_histogram(alpha = 0.7) +
  scale_fill_manual(values = c("ckd" = "blue", "notckd" = "green")) +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Count",
    fill = "CKD Classification"
  )
```

Histogram for serum creatinine vs classification of the disease

```{r}
ggplot(data_updated, aes(x = Serum.creatinine, fill = as.factor(Classification))) +
  geom_histogram(alpha = 0.7) +
  scale_fill_manual(values = c("ckd" = "brown", "notckd" = "yellow")) +
  labs(
    title = "Distribution of Serum.creatinine",
    x = "Serum.creatinine",
    y = "Count",
    fill = "CKD Classification"
  )
```

Age vs ckd/notckd - scatter plot
```{r}
ggplot(data_updated, aes(x = Age, y = as.factor(Classification), color = as.factor(Classification))) +
  geom_point() + geom_jitter(width = 0.1, height = 0.1) + 
  scale_color_manual(values = c("ckd" = "red", "notckd" = "blue"))
```

Box plots for age vs classification of ckd

```{r}
ggplot(data_updated, aes(x = as.factor(Classification), y = Age, color = as.factor(Classification))) +
  geom_boxplot() + scale_color_manual(values = c("ckd" = "red", "notckd" = "blue"))
  labs(x = "Classification", y = "Age")
```

Box plots for Serum.creatinine vs classification of ckd

```{r}
ggplot(data_updated, aes(x = as.factor(Classification), y = Serum.creatinine, color = as.factor(Classification))) +
  geom_boxplot() + scale_color_manual(values = c("ckd" = "purple", "notckd" = "blue"))
  labs(x = "Classification", y = "Age")
```

Pie chart showing the distribution of CKD classfication in the data set

```{r}
data_updated$classification <- as.factor(data_updated$Classification)
ckd_counts <- table(data_updated$Classification)

# Convert the table to a data frame for ggplot and calculate percentages
ckd_counts_df <- as.data.frame(ckd_counts) %>%
  mutate(Percentage = Freq / sum(Freq) * 100)

# Rename columns for clarity
names(ckd_counts_df) <- c("Classification", "Freq", "Percentage")

# Create the pie chart with percentage labels
ggplot(ckd_counts_df, aes(x = "", y = Freq, fill = Classification)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +  # This converts the bar chart to a pie chart
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), position = position_stack(vjust = 0.5)) +
  labs(
    title = "Distribution of CKD Classification",
    x = NULL,
    y = NULL,
    fill = "CKD Classification"
  ) +
  theme_void() +  # This removes the background, gridlines, and axes
  theme(legend.position = "right") 
```
Shapiro-Wilk test to check normality

```{r}
# Convert categorical variables to factor
data_updated <- data_updated %>%
  mutate_if(is.character, as.factor)

# Convert factor variables to numeric
data_updated <- data_updated %>%
  mutate_if(is.factor, as.numeric)

# Perform Shapiro-Wilk normality test on numeric variables
for (col in names(data_updated)[sapply(data_updated, is.numeric)]) {
  print(paste("Normality test for:", col))
  print(shapiro.test(data_updated[[col]]))
  print("---")
}
```
QQ plot to check the distribution of the data

```{r}
library(car)

# Loop through each numeric column in the dataset
par(mfrow = c(2, 2)) # Adjust the layout depending on the number of numeric variables
for (var in names(data_updated)) {
  if (is.numeric(data_updated[[var]])) {
    qqPlot(data_updated[[var]], main = paste("QQ Plot of", var))
  }
}
```

Mann Whitney U  test to check the significance for numerical variables
```{r}
wilcox.test(Age ~ Classification, data = data_updated)
wilcox.test(Blood.Pressure ~ Classification, data = data_updated)
wilcox.test(Albumin ~ Classification, data = data_updated)
wilcox.test(Serum.creatinine ~ Classification, data = data_updated)
wilcox.test(Hemoglobin ~ Classification, data = data_updated)
wilcox.test(Specific.Gravity ~ Classification, data = data_updated)
wilcox.test( Sugar~ Classification, data = data_updated)
wilcox.test(Red.Blood.Cells ~ Classification, data = data_updated)
wilcox.test(Pus.cells ~ Classification, data = data_updated)
wilcox.test(Pus.cell.clumps ~ Classification, data = data_updated)
wilcox.test(Bacteria ~ Classification, data = data_updated)
wilcox.test(Blood.glucose.random ~ Classification, data = data_updated)
wilcox.test(Blood.urea ~ Classification, data = data_updated)
wilcox.test(Sodium ~ Classification, data = data_updated)
wilcox.test(Potassium ~ Classification, data = data_updated)
wilcox.test(Packed.cell.volume ~ Classification, data = data_updated)
wilcox.test(White.blood.cell.count ~ Classification, data = data_updated)
wilcox.test(Red.blood.cell.count ~ Classification, data = data_updated)
```

Spearman Rank correlation test for  assessing the association between 2 numerical variables

```{r}
spearman_result <- cor.test(data_updated$Age, data_updated$Hemoglobin, method = "spearman", use = "complete.obs")
spearman_result

```

```{r}
spearman_result <- cor.test(data_updated$Age, data_updated$hypertension, method = "spearman", use = "complete.obs")
spearman_result
```

```{r}
spearman_result <- cor.test(data_updated$Age, data_updated$Albumin, method = "spearman", use = "complete.obs")
spearman_result
```

```{r}
install.packages("reshape2")
library(reshape2)
```
Correlation Heat map to check the association between 2 numerical variables
```{r}
# Install and load necessary packages
if (!require(reshape2)) install.packages("reshape2")
if (!require(ggplot2)) install.packages("ggplot2")
library(reshape2)
library(ggplot2)

cor_matrix <- cor(data_updated[, sapply(data_updated, is.numeric)], use = "pairwise.complete.obs")

# Melt the correlation matrix for ggplot2
melted_cor_matrix <- melt(cor_matrix)

# Create a heatmap with ggplot2 and add text annotations
ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +  # Tiles with white borders
  geom_text(aes(label = sprintf("%.2f", value)), color = "black", size = 0.8) +  # Add text labels
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Correlation") +
  theme(legend.key.height = unit(1, "in")) 
```

Chi square test for categorical variables
```{r}
library(stats)

# Convert categorical variables to factors
data_updated <- data_updated %>%
  mutate_if(is.character, as.factor)

# Perform Chi-Square tests
for (col in c("hypertension", "Diabetes.Mellitus", "Coronary.artery.disease", "Appetite", "Pedal.edema", "Anemia")) {
  print(paste("Chi-Square test for:", col))
  print(chisq.test(data_updated[[col]], data_updated$classification))
  print("---")
}
```
Logistic regression model for all the parameters
```{r}
library(readr)
#data <- data %>%
  #mutate_if(is.numeric, ~if_else(is.na(.), median(., na.rm = TRUE), .)) %>%
  #mutate_if(is.character, ~if_else(is.na(.), Mode(.), .)) # Define Mode function if not defined

# Convert categorical variables to factors
categorical_vars <- c("Red.Blood.Cells", "Pus.cells", "Pus.cell.clumps", "Bacteria", "hypertension", 
                      "Diabetes.Mellitus", "Coronary.artery.disease", "Appetite", "Pedal.edema", "Anemia")
data[categorical_vars] <- lapply(data[categorical_vars], as.factor)

# Ensure the target variable is a factor
data$Classification <- factor(data$Classification, levels = c("notckd", "ckd"))

# Fit the logistic regression model
# Including significant continuous and categorical variables identified from Wilcoxon and Chi-square tests
model <- glm(Classification ~ Age + `Blood.Pressure` + Albumin + `Serum.creatinine` + Hemoglobin +
             `Specific.Gravity` + Sugar + `Blood.glucose.random` + `Blood.urea` + Sodium + 
             `Packed.cell.volume` + `White.blood.cell.count` + `Red.blood.cell.count` + 
             `Red.Blood.Cells` + `Pus.cells` + `Pus.cell.clumps` + Bacteria + hypertension + 
             `Diabetes.Mellitus` + `Coronary.artery.disease` + Appetite + `Pedal.edema` + Anemia,
             data = data, family = binomial(link = "logit"))

# Summary of the model
summary(model)
```
VIF to check the multicollinearity of the fitted model
```{r}
# Load necessary libraries
library(car)
library(glmnet)

# Assuming 'data' is your dataset and 'Classification' is the binary outcome variable
# Fit a logistic regression model with all predictors
full_model <- glm(Classification ~ Age + Blood.Pressure + Albumin + Serum.creatinine + Hemoglobin +
                  Specific.Gravity + Sugar + Blood.glucose.random + Blood.urea +
                  Sodium + Packed.cell.volume + White.blood.cell.count + 
                  Red.blood.cell.count + hypertension + Diabetes.Mellitus + 
                  Coronary.artery.disease + Appetite + Pedal.edema + Anemia, 
                  data = data, family = binomial())

# Calculate VIF for the fitted model
vif_model <- vif(full_model)
print(vif_model[vif_model > 5])  # Print VIF values greater than 5
```
Logistic regression with few parameters
```{r}
# Fit the logistic regression model with selected variables
# Assuming the significant variables identified are 'Hemoglobin', 'Serum.creatinine', 'hypertension'
final_model <- glm(Classification ~ Hemoglobin + Serum.creatinine + Specific.Gravity, 
                   data = data, family = binomial())

# Display the summary of the final model
summary(final_model)
```
ROC curve and AUC values to evaluate model performance
```{r}

# Load packages
library(caret)
library(pROC)
library(e1071)

model <- glm(Classification ~ Hemoglobin + Serum.creatinine + hypertension,
             family = binomial(), data = data)
# Predict probabilities
probabilities <- predict(model, type = "response")

# Convert probabilities to binary outcomes based on a threshold of 0.5
predicted_classes <- ifelse(probabilities > 0.5, 1, 0)

# Actual classes
actual_classes <- as.numeric(data$Classification) - 1  # Adjust if necessary based on how your factor levels are coded

# Calculate confusion matrix
conf_matrix <- confusionMatrix(as.factor(predicted_classes), as.factor(actual_classes))
print(conf_matrix)

# Compute ROC curve and AUC
roc_result <- roc(actual_classes, probabilities)
auc_value <- auc(roc_result)
print(auc_value)

# Plot ROC curve
plot(roc_result, main="ROC Curve", col="#1c61b6")
abline(a=0, b=1, col="red", lty=2)  # Diagonal line for reference
```
Precision, Recall, F1 Score, and Accuracy
```{r}
# Precision, Recall, F1 Score, and Accuracy
precision <- conf_matrix$byClass["Pos Pred Value"]  # Precision (Positive Predictive Value)
recall <- conf_matrix$byClass["Sensitivity"]       # Recall (True Positive Rate)
f1_score <- 2 * (precision * recall) / (precision + recall)  # F1 Score
accuracy <- conf_matrix$overall["Accuracy"]         # Accuracy

# Print the metrics
print(paste("Precision:", precision))
print(paste("Recall:", recall))
print(paste("F1 Score:", f1_score))
print(paste("Accuracy:", accuracy))
```
